---
title: "AKFIN_WebService_Short"
author: "Jordan Watson/Matt Callahan"
date: "2/22/2021"
output: bookdown::html_document2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE)
```

# Goal: This document describes easy access to daily SST data averaged across common spatial strata for Alaska

## Introduction 
The AFSC and AKFIN are making satellite sea surface temperature (SST) data easier to access for commonly used spatial strata.    

Satellite SST data from NOAA Coral Reef Watch (https://coralreefwatch.noaa.gov/product/5km/index_5km_sst.php) are downloaded daily for the entire spatial extent of the EEZ around Alaska. The time series begins 1985-01-01 and extends through the present, with a 1-2 day latency period (data sourced from Coral Reef Watch except Jan-Mar 1985 which were downloaded directly from NESDIS). The raw data (i.e., full gridded dataset) are available to users with an AKFIN database account and can be easily queried from Oracle, R, or Python.  

Alternatively, we describe here an easy data access tool that requires no accounts, no passwords, no VPN - just internet access and R. Daily SST data have been averaged withihn several dozen commonly used spatial strata, including each of the NMFS management areas, each of the ecosystem regions (ESRs), the BSIERP regions, and the ADF&G statistical areas.

The greatest convenience of this tool is that no files need to be locally stored, nor do time series need to be updated and appended to existing data files. You will need R packages **httr** to pull data from a URL. Additional packages **tidyverse** and **lubridate** are used here to demonstrate plotting and manipulation but the object retrieved using **httr** can easily be manipulated using base R instead.

```{r}
library(httr) # For pulling data via a URL
library(tidyverse) # Data manipulation
library(lubridate) # Date formatting
library(sp) # For maps
library(rgdal) # For maps
library(ggrepel) #For maps
require(gridExtra) #For maps
```

```{r echo=FALSE, message=FALSE, results=FALSE, fig.cap="Map of four commonly used spatial strata for which SST data are averaged daily."}
fgdb <- "Data/Alaska_Marine_Management_Areas.gdb"
ogrListLayers(fgdb)
area<- readOGR(dsn=fgdb,layer="Alaska_Marine_Areas_dd")
test.df <- merge(fortify(area), as.data.frame(area), by.x="id", by.y=0) %>%
  mutate(long2=ifelse(long>0,long-360, long))

P1<-ggplot(data = filter(test.df, !is.na(NMFS_REP_AREA)), aes(x=long2, y=lat, group = group, lines=NMFS_REP_AREA)) +
  geom_polygon(fill=NA, color="black") +
  xlab("") + 
  ggtitle("NMFS areas")+
  theme_bw() +
  theme(legend.position="none",
         plot.title = element_text(hjust = 0.5))

P2<-ggplot(data = filter(test.df, !is.na(Ecosystem_Subarea)), aes(x=long2, y=lat, group = group, lines=Ecosystem_Subarea)) +
  geom_polygon(fill=NA, color="black") +
  xlab("") +
  ggtitle("Ecosystem subareas") +
  ylab("") +
    theme_bw() +
  theme(legend.position="none",
         plot.title = element_text(hjust = 0.5))

P3<-ggplot(data = filter(test.df, !is.na(STAT_AREA)), aes(x=long2, y=lat, group = group, lines=STAT_AREA)) +
  geom_polygon(fill=NA, color="black") +
  xlab("long") +
  ggtitle("Stat. areas") +
  theme_bw() +
  theme(legend.position="none",
        plot.title = element_text(hjust = 0.5))

P4<-ggplot(data = filter(test.df, BSIERP_ID>0), aes(x=long2, y=lat, group = group, lines=as.character(BSIERP_ID))) +
  geom_polygon(fill=NA, color="black") +
  xlab("long") +
  ggtitle("BSIERP regions") +
  ylab("") +
  theme_bw() +
  theme(legend.position="none",
        plot.title = element_text(hjust = 0.5))
grid.arrange(grobs = list(P1, P2, P3, P4))
```

Data extraction using this AKFIN web service is as simple as the statement below, which will query the time series of daily temperatures for NMFS are 640. The data can be saved as an object for manipulation or piped directly into downstream functions. Note that if you prefer base R instead of the tidyverse, you will still need to load **dplyr** for the bind_rows() function to work. Alternatively, you'll need a base R solution to reformat the input data.  

```{r}
head(httr::content(httr::GET('https://apex.psmfc.org/akfin/data_marts/akmp/nmfs_area_crw_avg_sst?nmfs_area=640&start_date=19850101&end_date=20220101'), type = "application/json") %>% 
  bind_rows)
```

With this tool, users can easily incorporate SST data into stock assessments and other processes. For example, one could plot a time series of average summer SST for NMFS areas 640 and 650 (Fig 2).   

```{r, message=FALSE, fig.cap="June-August SST 1985-2020 in EGOA NMFS areas."}
httr::content(httr::GET('https://apex.psmfc.org/akfin/data_marts/akmp/nmfs_area_crw_avg_sst?nmfs_area=640,650&start_date=19850101&end_date=20220101'), type = "application/json") %>% 
  bind_rows %>% 
  mutate(MONTH=month(as_date(READ_DATE))) %>% # Extract month from the date
  filter(MONTH==6 | MONTH==7 | MONTH==8) %>% # Filter summer months
  group_by(YEAR,NMFSAREA)%>%
    summarize(SST=mean(MEANSST))%>% # Average the data by year and area.
  ggplot(aes(as.numeric(YEAR),SST)) + 
  geom_line() + 
  facet_wrap(~NMFSAREA, nrow=2) +
  xlab("Year") +
  theme_bw()

```

## Web Service-How It Works:
AKFIN has created a set of web services or web APIs to streamline access to the average daily temperatures within each of the relevant strata. In the database backend at AKFIN, satellite SST files are downloaded daily as two separate netCDF files - one for positive and one for negative longitudes. The daily data are then clipped by merging them with a spatial lookup table that encompasses the exclusive economic zone (EEZ) of Alaska and merged into a single daily gridded dataset that includes 212,813 temperature records for each day. 

```{r cars}
#View strata included in the lookup table
lkp <- readRDS("Data/crwsst_spatial_lookup_table.RDS") 
str(lkp)
```

In the first example we will query SST within a NMFS area.  Currently, daily mean SST for each NMFS area can be queried individually. For the Bering Sea and Gulf of Alaska, the query filters only data where water depth is between 10 and 200m. For the Aleutian Islands, a depth filter is not implemented. Analysts that are interested in data for different depth ranges, custom spatial bounds, or aggregated NMFS areas can contact the authors and we will arrange for your request.

The web service enables us to query data using a URL, where the URL itself contains information for querying data from the server. One could paste the URL below into a browser and view the output there if desired. Below, we query the URL "https://apex.psmfc.org/akfin/data_marts/akmp/nmfs_area_crw_avg_sst?nmfs_area=640", where "nmfs_area_crw_avg_sst?" is the name of the dataset we are querying. This is the daily CRW SST dataset, which has been averaged by the field **nmfs_area**. A "?" separates the dataset name from the query criteria. The default behavior is to pull a single record, the most recent in the data set. The web service encodes the data as "json" so R must be told how to decode the data using the type = "application/json" argument. 

```{r}
httr::content(httr::GET("https://apex.psmfc.org/akfin/data_marts/akmp/nmfs_area_crw_avg_sst?nmfs_area=640"), type = "application/json")%>% 
  bind_rows
#By default the URL returns a list. We use "%>% bind_rows" here to convert to a data frame (tibble)
```
 
## Time Periods:
It is unlikely that you want a single temperature record. To instead query a time period, the user can specify a new time parameter to the URL with an "&" delimiter. Users can query a time period using "start_date" and "end_date", "read_date", or "dates_back" parameters. Note that date is sufficient, it is not necessary to specify a time component in the query. 


Most users will want to query the entire duration of the time series, which for the CRW SST dataset begins on 1985-04-01. To query the entire time series, specify "start_date" & "end_date". "end_date" must be included, but if you do not know the most recent date of the time series, you can choose an end date some time far in the future and it will query all of the data that exist without an error.  

```{r}
data <- httr::content(httr::GET('https://apex.psmfc.org/akfin/data_marts/akmp/nmfs_area_crw_avg_sst?nmfs_area=640&start_date=19850101&end_date=20220101'), type = "application/json") %>% 
  bind_rows

head(data)
```

The full time series yields more than 13,000 rows of data per area (i.e., daily data from 1985-01-01 to present).   
```{r}
dim(data)
```

Any time range can be chosen with "start_date" and "end_date", for example if one wanted to pull SST in NMFS area 640 in 1987.  

```{r, message=FALSE, fig.cap="SST for 1987 in NMFS area 640."}
httr::content(httr::GET('https://apex.psmfc.org/akfin/data_marts/akmp/nmfs_area_crw_avg_sst?nmfs_area=640&start_date=19870101&end_date=19880101'), type = "application/json") %>% 
  bind_rows %>% 
  mutate(date=as_date(READ_DATE)) %>% 
  ggplot(aes(date,MEANSST)) + 
  geom_line()+
  theme_bw()
```

You can query a specific date with "read_date". For example SST in MFS 640 on Y2K.  

```{r}
#Query the day after your date of interest because omitting the time component in read_date misses that day's reading.
httr::content(httr::GET('https://apex.psmfc.org/akfin/data_marts/akmp/nmfs_area_crw_avg_sst?nmfs_area=640&read_date=20000101'), type = "application/json") %>% 
  bind_rows 
```

You can specify a number of days prior to any date using a "days_back" parameter specification. For example the three days before Y2K.  

```{r}
httr::content(httr::GET('https://apex.psmfc.org/akfin/data_marts/akmp/nmfs_area_crw_avg_sst?nmfs_area=640&read_date=20000101&days_back=2'), type = "application/json") %>% 
  bind_rows 
```

If "read_date" is not specified, "days_back" returns the most recent SSTs. Here are SSTs for the last three days in NMFS 640.  

```{r}
httr::content(httr::GET('https://apex.psmfc.org/akfin/data_marts/akmp/nmfs_area_crw_avg_sst?nmfs_area=640&days_back=2'), type = "application/json") %>% 
  bind_rows 
```

## Spatial Extents  
To query multiple areas, separate the values by a comma. For example to query NMFS areas 640 and 650 (Southeast Alaska outside waters).  

```{r}
httr::content(httr::GET("https://apex.psmfc.org/akfin/data_marts/akmp/nmfs_area_crw_avg_sst?nmfs_area=640,650"), type = "application/json")%>% 
  bind_rows
```

The **Ecosystem_sub** field in our lookup table contains all of the different ecosystem regions that can be queried from the data. Points outside these ecosystem regions are listed as NA for this field.  

```{r}
unique(lkp$Ecosystem_sub)
```

To query the data for the "Southeastern Bering Sea", for example, add "ecosystem_sub=Southeastern%20Bering%20Sea", where spaces are filled by "%20".  

```{r}
httr::content(httr::GET('https://apex.psmfc.org/akfin/data_marts/akmp/ecosystem_sub_crw_avg_sst?ecosystem_sub=Southeastern%20Bering%20Sea'), type = "application/json") %>%
  bind_rows
```

Putting the pieces all together, the data can be queried directly from AKFIN and saved, manipulated, or visualized directly. Here we query the full time series for the Eastern Gulf of Alaska and for the Eastern Aleutian Islands.    

```{r, message=FALSE, fig.cap="Eastern GOA and Eastern Aleutians SST from 1985 - Present."}
httr::content(httr::GET('https://apex.psmfc.org/akfin/data_marts/akmp/ecosystem_sub_crw_avg_sst?ecosystem_sub=Eastern%20Gulf%20of%20Alaska,Eastern%20Aleutians&start_date=19850101&end_date=20220101'), type = "application/json") %>% 
  bind_rows %>% 
  mutate(date=as_date(READ_DATE)) %>% 
  ggplot(aes(date,MEANSST)) + 
  geom_line() + 
  facet_wrap(~ECOSYSTEM_SUB)
```

One could query and summarize data by month (week, year, etc) by simply grouping and summarizing the data in-line. Note that because the most recent year will be incomplete, its annual mean will be wonky so you would likely want to filter this out before plotting.    

```{r, message=FALSE, fig.cap="Annual average SST for NMFS areas 640 and 650."}
#Note that year is a character so it needs to be converted to an integer for continuous plotting.
httr::content(httr::GET('https://apex.psmfc.org/akfin/data_marts/akmp/nmfs_area_crw_avg_sst?nmfs_area=640,650&start_date=19850101&end_date=20220101'), type = "application/json") %>% 
  bind_rows %>% 
  mutate(date=as_date(READ_DATE),
         YEAR=as.numeric(YEAR)) %>%  
  group_by(YEAR,NMFSAREA) %>% 
  summarise(meansst=mean(MEANSST)) %>% 
  ggplot(aes(YEAR,meansst)) + 
  geom_line() + 
  geom_smooth() +
  facet_wrap(~NMFSAREA, nrow=2)+
  theme_bw()
```

## Marine Heatwave Bonus Material

Once you have daily time series for a spatial stratum it's really straightforward to use the heatwaveR package to characterize marine heatwaves (MHWs) in your data. The package vignette is fantastic (https://robwschlegel.github.io/heatwaveR/). We'll demonstrate a few quick examples below using our time series for NMFS area 640 (Eastern GOA). To keep this document short we only demonstrate plotting a single stratum but please contact us for additional code examples with multiple spatial strata if you are interested. We're happy to help with any indicator development.  

In this example we save the SST data as an object in this first code chunk but you could just as easily embed the data pull into the MHW code chunk that follows and never save this initial object to your workspace.  

```{r}
# Save the SST data to an object
updateddata <- httr::content(httr::GET('https://apex.psmfc.org/akfin/data_marts/akmp/nmfs_area_crw_avg_sst?nmfs_area=640&start_date=19850101&end_date=20211231'), type = "application/json") %>% 
  bind_rows %>% 
  mutate(date=as_date(READ_DATE)) %>% 
  data.frame %>% 
  dplyr::select(date,meansst=MEANSST,NMFSAREA) #I just simplify my data frame
```

The detect_event() and the ts2clm() functions from **heatwaveR** will analyze your dataset for marine heatwave status given a particular baseline period. We specify the earliest 30-years as a baseline by convention. The ts2clm() function generates marine heatwave thresholds from your baseline period. The detect_event() function creates a list of 2 data frames. The first, **climatology**, adds a series of columns to your data frame, including the seasonal climatology and MHW thresholds as well as flags for whether or not a record falls within a MHW. The second list object, **event**, includes summary information about each of the MHW events (e.g., max intensity, duration, start and end dates, etc.).  

```{r}
library(heatwaveR)
mhw <- detect_event(ts2clm(updateddata %>% 
                              rename(t=date,temp=meansst) %>% 
                              arrange(t), climatologyPeriod = c("1985-04-01", "2015-03-31"))) #Specify the baseline period.

str(mhw)
```

To create the common "flame" plots for MHWs, simply extract the climatology object from the list and plot it. Below we will filter the data since 2018-12-01 for NMFS area 640. The plotting code here is just the default code from the heatwaveR vignettes. You can easily see how to customize the aesthetics in normal tidy fashion.  

```{r}
mhw_clim <- mhw$climatology %>% filter(t>=as.Date("2018-12-01"))  #Extract the MHW data

ggplot(data = mhw_clim, aes(x = t)) +
  geom_flame(aes(y = temp, y2 = thresh, fill = "all"), show.legend = T) +
  geom_flame(data = mhw_clim, aes(y = temp, y2 = thresh, fill = "top"),  show.legend = T) +
  geom_line(aes(y = temp, colour = "temp")) +
  geom_line(aes(y = thresh, colour = "thresh"), size = 1.0) +
  geom_line(aes(y = seas, colour = "seas"), size = 1.2) +
  scale_colour_manual(name = "Line Colour",
                      values = c("temp" = "black", 
                                 "thresh" =  "forestgreen", 
                                 "seas" = "grey80")) +
  scale_fill_manual(name = "Event Colour", 
                    values = c("all" = "salmon", 
                               "top" = "red")) +
  scale_x_date(date_labels = "%b %Y") +
  guides(colour = guide_legend(override.aes = list(fill = NA))) +
  labs(y = expression(paste("Temperature [", degree, "C]")), x = NULL) + 
  theme(legend.position="top")
```

You can also tweak the image slightly to add the intensity categories of the MHWs. Again, the plotting code is verbatim from the heatwaveR vignettes. To illustrate the categories a bit better here we switch data sets to the Northern Bering Sea ecosystem region. We have also consolidated some of the code to reduce the number of intermediate objects and we filter the data from 2019-01-01 to 2019-12-31.  

```{r}
#  Here we'll use an example where we do not save the SST data as a separate object first, simply embedding it into the ts2clm() function
clim_cat <- (detect_event(ts2clm(
  httr::content(httr::GET('https://apex.psmfc.org/akfin/data_marts/akmp/ecosystem_sub_crw_avg_sst?ecosystem_sub=Northern%20Bering%20Sea&start_date=19850101&end_date=20211231'), type = "application/json") %>% 
                                   bind_rows %>% 
                                   mutate(date=as_date(READ_DATE)) %>% 
                                   data.frame %>% 
                                   dplyr::select(t=date,temp=MEANSST) %>% 
                                   arrange(t), climatologyPeriod = c("1985-04-01", "2015-03-31"))))$climatology %>% #Specify the baseline period.
  dplyr::mutate(diff = thresh - seas,
                thresh_2x = thresh + diff,
                thresh_3x = thresh_2x + diff,
                thresh_4x = thresh_3x + diff) %>% 
  filter(t>=as.Date("2019-01-01") & t<=as.Date("2019-12-31")) # Select the time period to display.

# Set line colours
lineColCat <- c(
  "Temperature" = "black",
  "Climatology" = "gray20",
  "Threshold" = "darkgreen",
  "2x Threshold" = "darkgreen",
  "3x Threshold" = "darkgreen",
  "4x Threshold" = "darkgreen"
  )

# Set category fill colours
fillColCat <- c(
  "Moderate" = "#ffc866",
  "Strong" = "#ff6900",
  "Severe" = "#9e0000",
  "Extreme" = "#2d0000"
  )


ggplot(data = clim_cat, aes(x = t, y = temp)) +
  geom_flame(aes(y2 = thresh, fill = "Moderate")) +
  geom_flame(aes(y2 = thresh_2x, fill = "Strong")) +
  geom_flame(aes(y2 = thresh_3x, fill = "Severe")) +
  geom_flame(aes(y2 = thresh_4x, fill = "Extreme")) +
  geom_line(aes(y = thresh_2x, col = "2x Threshold"), size = 0.7, linetype = "dashed") +
  geom_line(aes(y = thresh_3x, col = "3x Threshold"), size = 0.7, linetype = "dotdash") +
  geom_line(aes(y = thresh_4x, col = "4x Threshold"), size = 0.7, linetype = "dotted") +
  geom_line(aes(y = seas, col = "Climatology"), size = 0.7) +
  geom_line(aes(y = thresh, col = "Threshold"), size = 0.7) +
  geom_line(aes(y = temp, col = "Temperature"), size = 0.6) +
  scale_colour_manual(name = NULL, values = lineColCat,
                      breaks = c("Temperature", "Climatology", "Threshold",
                                 "2x Threshold", "3x Threshold", "4x Threshold")) +
  scale_fill_manual(name = NULL, values = fillColCat, guide = FALSE) +
  scale_x_date(date_labels = "%b %Y") +
  guides(colour = guide_legend(override.aes = list(linetype = c("solid", "solid", "solid",
                                                                "dashed", "dotdash", "dotted"),
                                                   size = c(0.6, 0.7, 0.7, 0.7, 0.7, 0.7)))) +
  labs(y = "Temperature [°C]", x = NULL) +
  theme(legend.position="top")
```


References
Skirving, W; Marsh, B; De La Cour, J; Liu, G; Harris, A; Maturi, E; Geiger, E; Eakin, CM. CoralTemp and the Coral Reef Watch Coral Bleaching Heat Stress Product Suite Version 3.1. Remote Sens. 2020, 12, 3856; https://doi.org/10.3390/rs12233856.

