---
output:
  word_document:
    pandoc_args: ["--metadata-file=header.yaml"]
    reference_docx: styles_reference.docx
    df_print: kable
csl: "../cite/citestyle.csl"
bibliography: "../cite/bibliography.bib"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, error = FALSE, message = FALSE)
```

# Abstract

The proliferation of operational satellite data has facilitated downstream data products catered towards specific fisheries applications in near real-time. We utilized such data accessibility to connect a suite of fishery-dependent data with spatially-explicit environmental information in the backend of the Alaska Region database environment at the Alaska Fisheries Information Network (AKFIN). For example, sea surface temperature (SST) data were linked to all fish tickets and observer in the Oracle backend from 2002 - present (more than one million records), and new data are automatically matched each day. We further extended the utility of satellite data products through customized spatial clipping of gridded satellite data extents to regions of interest for Alaska fisheries management. Full gridded data sets apportioned to Alaska management and research shapefile polygons can be queried from the AKFIN database. Alternatively, aggregated data products (e.g., time series of SST for individual NMFS regions or ecosystem areas) can be accessed via custom web services, or URL-based data queries. We demonstrate several queries of the web service and illustrate how this product can yield seamless integration with downstream analyses by detecting marine heatwaves in the Eastern Bering Sea ecosystem region. 

```{r}

```

